---
title: "Example datasets"
author: Helena L. Crowell
format:
  html:
    embed-resources: true
    self_contained: true
    code-folding: show
    toc-float: true
    toc-depth: 2
    toc: true
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    cache=TRUE,
    message=FALSE, 
    warning=FALSE, 
    fig.align="center")
```

# Preamble

```{r deps}
library(ggplot2)
library(VisiumIO)
library(patchwork)
library(OSTA.data)
library(DropletUtils)
library(SpatialExperiment)
library(SpatialExperimentIO)
```

```{r help}
#| code-fold: true
# helper of viz.
pal <- c(
    "#e6194b", "#3cb44b", "#ffe119", "#4363d8", "#f58231", "#911eb4", "#42d4f4", 
    "#f032e6", "#bfef45", "#fabed4", "#469990", "#dcbeff", "#9a6324", "#fffac8", 
    "#800000", "#aaffc3", "#808000", "#ffd8b1", "#000075", "#a9a9a9", "#000000")
qal <- c(
    "#00007F", "#0000A9", "#0000D4", "#0000FF", "#002AFF",
    "#0054FF", "#007FFF", "#00A9FF", "#00D4FF", "#00FFFF",
    "#2AFFD4", "#54FFA9", "#7FFF7F", "#A9FF54", "#D4FF2A",
    "#FFFF00", "#FFD400", "#FFA900", "#FF7F00", "#FF5400",
    "#FF2A00", "#FF0000", "#D40000", "#A90000", "#7F0000")
plt <- \(o, i, s=2/3) {
    spatialCoordsNames(o) <- c("x", "y")
    df <- data.frame(colData(o), spatialCoords(o))
    if (i %in% rownames(o)) df[[i]] <- log1p(assay(o)[i, ])
    th <- if (is.numeric(df[[i]])) {
        list(
            theme(legend.key.width=unit(0.4, "lines")),
            theme(legend.key.height=unit(0.8, "lines")),
            scale_color_gradientn(colors=qal, na.value="grey90"))
    } else list(
        theme(legend.key.size=unit(0, "lines")),
        scale_color_manual(values=pal, na.value="grey90"),
        guides(col=guide_legend(override.aes=list(size=2))))
    ggplot(df, aes(x, y, col=.data[[i]])) +
        coord_equal() + theme_void() + th +
        geom_point(shape=16, stroke=0, size=s)
}
```

See also [OSTA](https://bioconductor.org/books/release/OSTA) for more details.

```{r}
# view available datasets
OSTA.data_list()
```

# [Janesick et al. (2023)](http://doi.org/10.1038/s41467-023-43458-x)

::: {.panel-tabset}

## Chromium

```{r jan-chr}
# retrieval
id <- "Chromium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
h5 <- list.files(td, "h5$", full.names=TRUE)
(sce <- read10xCounts(h5))
```

## Xenium

```{r jan-xen}
# retrieval
id <- "Xenium_HumanBreast1_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
(spe <- readXeniumSXE(td))
```

## Visium

```{r jan-vis}
# retrieval
id <- "Visium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
obj <- TENxVisium(
    spacerangerOut=td, 
    images="lowres", 
    format="h5")
spe <- import(obj)
# annotations
df <- read.csv(file.path(td, "annotation.csv"))
cs <- match(colnames(spe), df$Barcode)
spe$anno <- factor(df$Annotation[cs])
```

```{r vis-plt, fig.width=6, fig.height=3}
plt(spe, "anno", s=4/3)
```

:::

# [Oliveira et al. (2024)](http://doi.org/10.1101/2024.06.04.597233)

::: {.panel-tabset}

## Xenium

```{r oli-xen}
# retrieval
id <- "Xenium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
(spe <- readXeniumSXE(td))
```

```{r xen-plt, fig.width=7, fig.height=4}
y <- spatialCoords(spe)[, 2]
i <- y > 1e3 & y < 5e3
plt(spe[, i], "CEACAM6")
```

## Visium

```{r oli-vis}
# retrieval
id <- "Visium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
obj <- TENxVisium(
    spacerangerOut=td, 
    images="lowres", 
    format="h5")
(spe <- import(obj))
```

## VisiumHD

```{r oli-vhd}
# retrieval
id <- "VisiumHD_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
obj <- TENxVisiumHD(
    spacerangerOut=td, 
    images="lowres", 
    format="h5")
(spe <- import(obj))
```

:::

# CosMx

```{r tx}
# whether or not to load transcripts
# (skipped here to reduce runtime)
tx <- FALSE
```

::: {.panel-tabset}

## 1k-plex

```{r cos-1k}
# retrieval
id <- "CosMx1k_MouseBrain1"
pa <- OSTA.data_load(id, mol=tx)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
(spe <- readCosmxSXE(td, addTx=tx))
```

```{r cos-plt, fig.width=7, fig.height=4}
plt(spe, "Mbp")
```

## 6k-plex

```{r cos-6k}
# retrieval
id <- "CosMx6k_HumanBrain"
pa <- OSTA.data_load(id, mol=tx)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# importing
(spe <- readCosmxSXE(td, addTx=tx))
```

## 18k-plex

- from [Crowell et al.](http://doi.org/10.1101/2025.06.23.660674)
- human colorectal cancer (CRC) sections
- low- and high-resolution cell type annotations
- histopathological annotations: healthy, pre- and malignant
- data is available as .rds (SCE), .h5ad, and `alabaster` artifacts

```{r cox-wtx}
# for other sections & data formats, see 
# https://zenodo.org/records/15574384

# retrieval
options(timeout=1e3)
url <- "https://zenodo.org/records/15574384/files/221.zip"
tf <- tempfile(fileext=".zip")
download.file(url, tf)
# decompression
pa <- dirname(tf)
unzip(tf, exdir=pa)
# importing
library(alabaster.sce)
nm <- file.path(pa, "221")
sce <- readObject(nm)
# coersion
xy <- sprintf("Center%s_global_mm", c("X", "Y"))
xy <- as.matrix(colData(sce)[xy])
(spe <- toSpatialExperiment(sce, spatialCoords=xy))
```

```{r wtx-plt, fig.width=10, fig.height=4}
plt(spe, "lv1", s=1/5) + scale_color_manual(
    values=c("gold", "turquoise", "deeppink")) +
plt(spe, "typ", s=1/5) + scale_color_manual(
    values=c("tomato", "seagreen", "dodgerblue", "royalblue", "slateblue")) 
```

## tonsil

- from [Crowell et al. (2025)](http://dx.doi.org/10.1002/eji.70121)
- 3 human tonsil sections (1 child, 2 adult)
- low- and high-resolution cell type annotations
- annotated regions of interest (ROIs): germinal centers, epithelium
- ROIs were identifier programmatically using `r BiocStyle::Biocpkg("sosta")`

```{r cos-slo}
# for other sections & data formats, see 
# https://zenodo.org/records/16782547

# retrieval
options(timeout=1e3)
url <- "https://zenodo.org/records/16782547/files/C1.zip"
tf <- tempfile(fileext=".zip")
download.file(url, tf)
# decompression
pa <- dirname(tf)
unzip(tf, exdir=pa)
# importing
library(alabaster.sce)
nm <- file.path(pa, "C1")
sce <- readObject(nm)
# coersion
xy <- sprintf("Center%s_global_mm", c("X", "Y"))
xy <- as.matrix(colData(sce)[xy])
(spe <- toSpatialExperiment(sce, spatialCoords=xy))
```

```{r plt-slo, fig.width=12, fig.height=8}
roi <- as.character(spe$gcs); roi[.] <- as.character(spe$epi[. <- !is.na(spe$epi)])
spe$roi <- roi; col <- colorRampPalette(pal)(length(unique(spe$roi)))
plt(spe, "lv1") + plt(spe, "roi") + 
    theme(legend.position="none") +
    scale_color_manual(values=col, na.value="grey95") 
```

:::

# MERFISH

- from [Cadinu et al. (2024)](https://doi.org/10.5061/dryad.rjdfn2zh3)
- 943 genes x 1.35 million cells
- 4 timepoints (healthy + 3/9/21 days post-onset of colitis)

```{r mer}
library(MerfishData)
(spe <- MouseColonIbdCadinu2024())
```

```{r mer-plt, fig.width=6, fig.height=4}
ids <- with(
    colData(spe), paste(sep=",", 
    mouse_id, slice_id, technical_repeat_number))
plt(spe[, ids == ids[1]], "tier1")
```

::: {.callout-note icon=false, collapse=true}

# Session

```{r si}
sessionInfo()
```

:::