---
title: "Compute mean intensity per polygon"
format: html
author: Davide Risso
---

# Setup

```{r setup}
library(Rarr)
library(ZarrArray)
library(sf)
library(ggplot2)
library(duckspatial)

sf::sf_use_s2(FALSE)
```

# Read files

## Read Zarr

```{r readin}
zarr_file <- "../data/TCGA-02-0001-01Z-00-DX1.83fce43e-42ac-4dcd-b156-2908e75f2e47.zarr/images/images/0"

A <- ZarrArray::ZarrArray(zarr_file)
is(A, "DelayedArray")
zarray_info <- Rarr::zarr_overview(zarr_file, as_data_frame = TRUE)
Rarr::zarr_overview(zarr_file)
```

## Read parquet

```{r parquet}
poly_crs <- 4326   # CRS for polygon coordinates (pixel space, no units)
img_shape_scale <- 1

# polys stays as a lazy duckspatial table — never fully materialised
polys <- duckspatial::ddbs_open_dataset(
    "../data/TCGA-02-0001-01Z-00-DX1.83fce43e-42ac-4dcd-b156-2908e75f2e47.parquet",
    crs = poly_crs
)
```

# Extract chunk information

Extract chunk geometry from the Zarr metadata.

```{r firstchunk}
img_shape   <- zarray_info$dim[[1]]      # e.g. [33776, 46030] or [3, 33776, 46030]
chunk_shape <- zarray_info$chunk_dim[[1]]     # e.g. [1024, 1024]   or [1, 1024, 1024]

# Determine which axes are spatial (Y and X). Adjust indices if multichannel.
# Assumption: last two axes are Y (rows) and X (cols).
n_dims  <- length(img_shape)
y_ax    <- n_dims - 1   # 1-based index for Rarr slice (n_dims - 1 + 1 = n_dims... see below)
x_ax    <- n_dims

img_h   <- img_shape[n_dims - 1]    # total height in pixels
img_w   <- img_shape[n_dims]        # total width  in pixels
chunk_h <- chunk_shape[n_dims - 1]
chunk_w <- chunk_shape[n_dims]
```

Build a grid of chunks.

```{r grid}
# Grid of chunk origins (1-based R indices, top-left corner of each chunk)
row_origins <- seq(1, img_h, by = chunk_h)
col_origins <- seq(1, img_w, by = chunk_w)
chunk_grid <- expand.grid(row0 = row_origins, col0 = col_origins)
```

## Focus on third chunk

First two chunks are empty (?), let's look at the third chunk.

```{r subset}
row0 <- chunk_grid$row0[3]
col0 <- chunk_grid$col0[3]

# Pixel bounding box of this chunk (1-based R indices, inclusive)
row1 <- min(row0 + chunk_h - 1, img_h)
col1 <- min(col0 + chunk_w - 1, img_w)

# Subset polygons for the third chunk.
image_roi <- c(xmin = col0 - 1, ymin = row0 - 1, 
               xmax = col1, ymax = row1)
shape_roi <- image_roi*img_shape_scale
chunk_bbox <- st_bbox(shape_roi, crs = st_crs(poly_crs))

chunk_sf <- st_sf(geom = st_as_sfc(chunk_bbox))

chunk_polys <- ddbs_filter(polys, chunk_sf) |> ddbs_collect()
chunk_polys
```

Compute per-polygon mean intensity for the third chunk. The image chunk is read once, then reused for both the RGB background and the per-channel extraction.

```{r third-chunk-mean-intensity}
plot_roi <- as.list(image_roi)

# Single image read for this chunk: [C, Y, X].
third_chunk <- as.array(A[1:3, row0:row1, col0:col1])
third_chunk_yxc <- base::aperm(third_chunk, c(2, 3, 1))

# Scale a separate view for RGB plotting without changing the statistic values.
third_chunk_rgb <- third_chunk_yxc
storage.mode(third_chunk_rgb) <- "double"

if (max(third_chunk_rgb, na.rm = TRUE) > 1) {
  third_chunk_rgb <- third_chunk_rgb / 255
}

third_chunk_raster <- as.raster(third_chunk_rgb)

point_in_polygon <- function(px, py, vx, vy) {
  inside <- rep(FALSE, length(px))
  j <- length(vx)

  for (i in seq_along(vx)) {
    crosses <- (vy[i] > py) != (vy[j] > py)
    if (any(crosses)) {
      x_cross <- (vx[j] - vx[i]) * (py[crosses] - vy[i]) /
        (vy[j] - vy[i]) + vx[i]
      inside[crosses] <- xor(inside[crosses], px[crosses] < x_cross)
    }
    j <- i
  }

  inside
}

polygon_channel_means <- function(poly, chunk, row0, row1, col0, col1) {
  bbox <- sf::st_bbox(poly)
  n_channels <- dim(chunk)[1]
  empty <- list(n_pixels = 0L, means = rep(NA_real_, n_channels))

  col_start <- max(col0, floor(bbox[["xmin"]]) + 1L)
  col_end <- min(col1, ceiling(bbox[["xmax"]]))
  row_start <- max(row0, floor(bbox[["ymin"]]) + 1L)
  row_end <- min(row1, ceiling(bbox[["ymax"]]))

  if (col_start > col_end || row_start > row_end) {
    return(empty)
  }

  cols <- seq.int(col_start, col_end)
  rows <- seq.int(row_start, row_end)

  centers <- expand.grid(
    col = cols,
    row = rows,
    KEEP.OUT.ATTRS = FALSE
  )
  px <- centers$col - 0.5
  py <- centers$row - 0.5

  coords <- as.data.frame(sf::st_coordinates(poly))
  part_id <- if ("L3" %in% names(coords)) coords$L2 else rep(1L, nrow(coords))
  ring_id <- if ("L1" %in% names(coords)) coords$L1 else rep(1L, nrow(coords))

  inside <- rep(FALSE, nrow(centers))
  for (part in unique(part_id)) {
    in_part <- rep(FALSE, nrow(centers))
    part_rows <- part_id == part
    rings <- unique(ring_id[part_rows])

    for (ring in rings) {
      ring_rows <- part_rows & ring_id == ring
      ring_inside <- point_in_polygon(
        px,
        py,
        coords$X[ring_rows],
        coords$Y[ring_rows]
      )

      if (ring == 1L) {
        in_part <- ring_inside
      } else {
        in_part <- in_part & !ring_inside
      }
    }

    inside <- inside | in_part
  }

  if (!any(inside)) {
    return(empty)
  }

  local_rows <- centers$row[inside] - row0 + 1L
  local_cols <- centers$col[inside] - col0 + 1L

  means <- vapply(seq_len(n_channels), function(channel) {
    vals <- chunk[cbind(rep.int(channel, length(local_rows)), local_rows, local_cols)]
    mean(vals, na.rm = TRUE)
  }, numeric(1))

  list(n_pixels = length(local_rows), means = means)
}

channel_cols <- paste0("channel_", seq_len(dim(third_chunk)[1]))
intensity_list <- lapply(seq_len(nrow(chunk_polys)), function(i) {
  stats <- polygon_channel_means(
    chunk_polys[i, ],
    third_chunk,
    row0,
    row1,
    col0,
    col1
  )
  out <- data.frame(
    polygon_id = i,
    n_pixels = stats$n_pixels,
    stringsAsFactors = FALSE
  )
  out[channel_cols] <- as.list(stats$means)
  out
})

chunk_intensity <- do.call(rbind, intensity_list)
chunk_intensity$mean_intensity <- rowMeans(
  chunk_intensity[, channel_cols, drop = FALSE],
  na.rm = TRUE
)
chunk_intensity$mean_intensity[chunk_intensity$n_pixels == 0L] <- NA_real_

# Match ggplot display coordinates: raster row 1 is drawn at the top, while
# segmentation y coordinates are image pixel coordinates increasing down.
chunk_polys_plot <- chunk_polys
sf::st_geometry(chunk_polys_plot) <-
  sf::st_geometry(chunk_polys_plot) *
  matrix(c(1, 0, 0, -1), 2, 2) +
  c(0, plot_roi$ymin + plot_roi$ymax)

chunk_polys_plot$polygon_id <- seq_len(nrow(chunk_polys_plot))
chunk_polys_intensity <- chunk_polys_plot
matched <- match(chunk_polys_intensity$polygon_id, chunk_intensity$polygon_id)
for (nm in c("n_pixels", channel_cols, "mean_intensity")) {
  chunk_polys_intensity[[nm]] <- chunk_intensity[[nm]][matched]
}

head(chunk_intensity)

ggplot() +
#  annotation_raster(
#    third_chunk_raster,
#    xmin = plot_roi$xmin,
#    xmax = plot_roi$xmax,
#    ymin = plot_roi$ymin,
#    ymax = plot_roi$ymax,
#    interpolate = FALSE
#  ) +
  geom_sf(
    data = chunk_polys_intensity,
    aes(fill = mean_intensity),
    color = "black",
    linewidth = 0.05,
    alpha = 0.75
  ) +
  scale_fill_viridis_c(
    option = "magma",
    na.value = "transparent",
    name = "Mean intensity"
  ) +
  coord_sf(
    xlim = c(plot_roi$xmin, plot_roi$xmax),
    ylim = c(plot_roi$ymin, plot_roi$ymax),
    expand = FALSE,
    datum = NA
  ) +
  theme_void() +
  labs(title = "Third Zarr Chunk: Polygons Colored by Mean Intensity")
```

# Iterate over all chunks

The same chunk-first strategy can be extended to the full image. Each Zarr
chunk is read at most once, polygons are restricted to the pixel centers inside
that chunk, and per-polygon sums are accumulated across chunks before the final
means are computed.

```{r all-chunks-mean-intensity}
rm(
  list = intersect(
    c("third_chunk", "third_chunk_yxc", "third_chunk_rgb", "third_chunk_raster"),
    ls()
  )
)
invisible(gc())

parquet_file <- "../data/TCGA-02-0001-01Z-00-DX1.83fce43e-42ac-4dcd-b156-2908e75f2e47.parquet"
channel_index <- seq_len(img_shape[1])
channel_cols <- paste0("channel_", channel_index)
n_channels <- length(channel_index)

if (!exists("point_in_polygon")) {
  point_in_polygon <- function(px, py, vx, vy) {
    inside <- rep(FALSE, length(px))
    j <- length(vx)

    for (i in seq_along(vx)) {
      crosses <- (vy[i] > py) != (vy[j] > py)
      if (any(crosses)) {
        x_cross <- (vx[j] - vx[i]) * (py[crosses] - vy[i]) /
          (vy[j] - vy[i]) + vx[i]
        inside[crosses] <- xor(inside[crosses], px[crosses] < x_cross)
      }
      j <- i
    }

    inside
  }
}

polygon_channel_sums <- function(poly, chunk, row0, row1, col0, col1) {
  bbox <- sf::st_bbox(poly)
  empty <- list(n_pixels = 0L, sums = rep(0, dim(chunk)[1]))

  col_start <- max(col0, floor(bbox[["xmin"]]) + 1L)
  col_end <- min(col1, ceiling(bbox[["xmax"]]))
  row_start <- max(row0, floor(bbox[["ymin"]]) + 1L)
  row_end <- min(row1, ceiling(bbox[["ymax"]]))

  if (col_start > col_end || row_start > row_end) {
    return(empty)
  }

  centers <- expand.grid(
    col = seq.int(col_start, col_end),
    row = seq.int(row_start, row_end),
    KEEP.OUT.ATTRS = FALSE
  )
  px <- centers$col - 0.5
  py <- centers$row - 0.5

  coords <- as.data.frame(sf::st_coordinates(poly))
  part_id <- if ("L3" %in% names(coords)) coords$L2 else rep(1L, nrow(coords))
  ring_id <- if ("L1" %in% names(coords)) coords$L1 else rep(1L, nrow(coords))

  inside <- rep(FALSE, nrow(centers))
  for (part in unique(part_id)) {
    in_part <- rep(FALSE, nrow(centers))
    part_rows <- part_id == part
    rings <- unique(ring_id[part_rows])

    for (ring in rings) {
      ring_rows <- part_rows & ring_id == ring
      ring_inside <- point_in_polygon(
        px,
        py,
        coords$X[ring_rows],
        coords$Y[ring_rows]
      )

      if (ring == 1L) {
        in_part <- ring_inside
      } else {
        in_part <- in_part & !ring_inside
      }
    }

    inside <- inside | in_part
  }

  if (!any(inside)) {
    return(empty)
  }

  local_rows <- centers$row[inside] - row0 + 1L
  local_cols <- centers$col[inside] - col0 + 1L

  sums <- vapply(seq_len(dim(chunk)[1]), function(channel) {
    vals <- chunk[cbind(rep.int(channel, length(local_rows)), local_rows, local_cols)]
    sum(vals, na.rm = TRUE)
  }, numeric(1))

  list(n_pixels = length(local_rows), sums = sums)
}

con <- DBI::dbConnect(duckdb::duckdb())
invisible(DBI::dbExecute(con, "LOAD spatial"))

invisible(DBI::dbExecute(
  con,
  paste0(
    "CREATE TEMP TABLE polygons_indexed AS ",
    "SELECT CAST(row_number() OVER () AS INTEGER) AS polygon_id, * ",
    "FROM read_parquet(", DBI::dbQuoteString(con, parquet_file), ")"
  )
))

n_polygons <- as.integer(DBI::dbGetQuery(
  con,
  "SELECT count(*) AS n FROM polygons_indexed"
)$n)

pixel_counts <- integer(n_polygons)
channel_sums <- matrix(
  0,
  nrow = n_polygons,
  ncol = n_channels,
  dimnames = list(NULL, channel_cols)
)

pb <- utils::txtProgressBar(min = 0, max = nrow(chunk_grid), style = 3)

for (chunk_i in seq_len(nrow(chunk_grid))) {
  utils::setTxtProgressBar(pb, chunk_i)

  row0_i <- chunk_grid$row0[chunk_i]
  col0_i <- chunk_grid$col0[chunk_i]
  row1_i <- min(row0_i + chunk_h - 1L, img_h)
  col1_i <- min(col0_i + chunk_w - 1L, img_w)

  chunk_roi <- c(
    xmin = col0_i - 1L,
    ymin = row0_i - 1L,
    xmax = col1_i,
    ymax = row1_i
  )
  chunk_wkt <- sf::st_as_text(sf::st_as_sfc(sf::st_bbox(chunk_roi)))

  chunk_sql <- paste0(
    "SELECT polygon_id, geometry ",
    "FROM polygons_indexed ",
    "WHERE ST_Intersects(ST_GeomFromText(",
    DBI::dbQuoteString(con, chunk_wkt),
    "), geometry)"
  )
  chunk_tbl <- DBI::dbGetQuery(con, chunk_sql)

  if (nrow(chunk_tbl) == 0L) {
    next
  }

  chunk_polys_i <- sf::st_sf(
    chunk_tbl["polygon_id"],
    geometry = sf::st_as_sfc(
      structure(chunk_tbl$geometry, class = "WKB"),
      EWKB = TRUE,
      crs = poly_crs
    )
  )

  # Read this Zarr chunk exactly once, then discard it before the next chunk.
  chunk_data <- as.array(A[channel_index, row0_i:row1_i, col0_i:col1_i])

  for (poly_i in seq_len(nrow(chunk_polys_i))) {
    polygon_id <- chunk_polys_i$polygon_id[poly_i]
    partial <- polygon_channel_sums(
      chunk_polys_i[poly_i, ],
      chunk_data,
      row0_i,
      row1_i,
      col0_i,
      col1_i
    )

    if (partial$n_pixels == 0L) {
      next
    }

    pixel_counts[polygon_id] <- pixel_counts[polygon_id] + partial$n_pixels
    channel_sums[polygon_id, ] <- channel_sums[polygon_id, ] + partial$sums
  }

  rm(chunk_data, chunk_polys_i, chunk_tbl)
  invisible(gc(FALSE))
}

close(pb)

has_pixels <- pixel_counts > 0L
all_chunk_intensity <- data.frame(
  polygon_id = which(has_pixels),
  n_pixels = pixel_counts[has_pixels],
  channel_sums[has_pixels, , drop = FALSE] /
    pixel_counts[has_pixels],
  row.names = NULL
)
all_chunk_intensity$mean_intensity <- rowMeans(
  all_chunk_intensity[, channel_cols, drop = FALSE],
  na.rm = TRUE
)

polygon_attrs <- DBI::dbGetQuery(
  con,
  paste(
    "SELECT polygon_id, type_id, type_prob, magnification, scale_factor,",
    "minx, miny, maxx, maxy",
    "FROM polygons_indexed"
  )
)

all_chunk_intensity <- merge(
  polygon_attrs,
  all_chunk_intensity,
  by = "polygon_id",
  all.y = TRUE
)

DBI::dbDisconnect(con, shutdown = TRUE)

head(all_chunk_intensity)
```

```{r plot_all}
ggplot(all_chunk_intensity, aes(minx, miny, color=channel_1)) +
  geom_point(size=0.1) +
  scale_color_viridis_c(
    option = "magma",
    na.value = "transparent",
    name = "Mean intensity"
  ) +
  theme_void()

ggplot(all_chunk_intensity, aes(minx, miny, color=channel_2)) +
  geom_point(size=0.1) +
  scale_color_viridis_c(
    option = "magma",
    na.value = "transparent",
    name = "Mean intensity"
  ) +
  theme_void()

ggplot(all_chunk_intensity, aes(minx, miny, color=channel_3)) +
  geom_point(size=0.1) +
  scale_color_viridis_c(
    option = "magma",
    na.value = "transparent",
    name = "Mean intensity"
  ) +
  theme_void()

library(terra)
ome.tiff.file <- "../data/TCGA-02-0001-01Z-00-DX1.83fce43e-42ac-4dcd-b156-2908e75f2e47.ome.tiff"
rasterized <- rast(ome.tiff.file)
plot(rasterized)

```