---
title: "Crop image"
format: html
author: "Davide Risso"
---

# Setup

```{r setup}
library(ImageArray)
library(RBioFormats)
library(ggplot2)
```

# Read tiff image

```{r tiff}
ome.tiff.file <- "../data/TCGA-02-0001-01Z-00-DX1.83fce43e-42ac-4dcd-b156-2908e75f2e47.ome.tiff"
read.metadata(ome.tiff.file)

imgarray <- createImageArray(ome.tiff.file, series = 1, resolution = 1)
imgarray
```

# Crop

```{r crop}
roi <- list(xmin = 5000, ymin = 17000, xmax = 6000, ymax = 18000)

imgarray_sub <- crop(imgarray, index = list(roi$xmin:roi$xmax, roi$ymin:roi$ymax, NULL))

img_raster <- as.raster(imgarray_sub, max.pixel.size = 100)

plot(img_raster)

ggplot(data.frame(x = 0, y = 0), aes(x, y)) +
  coord_fixed(expand = FALSE,
              xlim = c(0, dim(img_raster)[2]),
              ylim = c(0, dim(img_raster)[1])) +
  annotation_raster(img_raster, 
                    0, dim(img_raster)[2],
                    dim(img_raster)[1], 0, interpolate = FALSE)
```

Alternative way: terra.

```{r terra}
library(terra)

rasterized <- rast(ome.tiff.file)
plot(rasterized)

sub_img <- terra::crop(rasterized, ext(roi$xmin, roi$xmax, roi$ymin, roi$ymax))
sub_img

plot(sub_img)
```

# Read zarr image

Let's try to do the same but starting from a Zarr file.

```{r zarr}
library(ZarrArray)
zarr_file <- "../data/TCGA-02-0001-01Z-00-DX1.83fce43e-42ac-4dcd-b156-2908e75f2e47.zarr/images/images/0"

A <- ZarrArray(zarr_file)
is(A, "DelayedArray")

img_cropped <- A[1:3, (roi$ymin):(roi$ymax), (roi$xmin):(roi$xmax)]

img_array <- as.array(img_cropped)
img_array <- aperm(img_array, c(2, 3, 1))
sub_img2 <- rast(img_array)
RGB(sub_img2) <- 1:3

plot(sub_img2)
```

It looks mirrored in the y axis, but that's just a convention from the Python zarr file (I think).

# Read GeoParquet

We can use `duckspatial` to read the GeoParquet file and crop it to the same coordinates as the image. This is efficient as it only reads the polygons that intersect with the crop area.

```{r duckspatial}
library(duckdb)
library(duckspatial)
library(duckplyr)
library(sf)

parquet_file <- "../data/TCGA-02-0001-01Z-00-DX1.83fce43e-42ac-4dcd-b156-2908e75f2e47.parquet"

# Open dataset
ddbs <- ddbs_open_dataset(parquet_file)

# Define bounding box for cropping
stbbox <- st_bbox(c(xmin = roi$xmin, xmax = roi$xmax,
                    ymin = roi$ymin, ymax = roi$ymax))
wkb_string <- st_as_text(st_as_sfc(stbbox))

# Filter and load as sf object
polygons_sub <- ddbs |>
  filter(ST_Intersects(ST_GeomFromText(wkb_string), geometry)) |>
  st_as_sf()

polygons_sub

# Plot polygons
ggplot() +
  geom_sf(data = polygons_sub) +
  geom_sf(data = st_as_sfc(stbbox), fill = NA) +
  theme_minimal() +
  labs(title = "Cropped GeoParquet Polygons")
```

# Overlay Image and Polygons

```{r overlay}
# Ensure we only have 3 channels (RGB)
sub_img_3ch <- sub_img[[1:3]]

# Set RGB channels for base plot
RGB(sub_img_3ch) <- 1:3
plotRGB(sub_img_3ch)
plot(st_geometry(polygons_sub), add = TRUE, border = "white", lwd = 0.5)

# For ggplot2, we need a single-layer raster of hex colors.
# We manually convert to an array and scale to [0, 1] for as.raster()
img_arr <- as.array(sub_img_3ch)

# Scale values to [0, 1] if they aren't already
img_min <- min(img_arr, na.rm = TRUE)
img_max <- max(img_arr, na.rm = TRUE)
img_arr_scaled <- (img_arr - img_min) / (img_max - img_min)

# as.raster() on a (H, W, 3) array produces the correct hex-encoded matrix
sub_img_raster_obj <- as.raster(img_arr_scaled)

ggplot() +
  annotation_raster(sub_img_raster_obj, 
                    roi$xmin, roi$xmax,
                    roi$ymax, roi$ymin, interpolate = FALSE) +
  geom_sf(data = polygons_sub, fill = "white", color = "white", alpha = 0.5) +
  coord_sf(xlim = c(roi$xmin, roi$xmax), ylim = c(roi$ymin, roi$ymax), expand = FALSE) +
  theme_void() +
  labs(title = "Overlay: Terra Image and Segmentation")
```

